---
title: "Assignment 6 - QMM"
author: "Shahbaz"
date: "2025-10-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# STEP 1 - Loading Necessary Libraries
```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
if(!require(lpSolve)) install.packages("lpSolve", quiet = TRUE)
if(!require(knitr)) install.packages("knitr", quiet = TRUE)
library(lpSolve)
library(knitr)
```


# Problem Formulation

Understanding the Problem Data
Heart Start has two plants (A and B) with the following production capacities and unit costs:

Plant A: Capacity = 100 units, Production Cost = $600/unit

Plant B: Capacity = 120 units, Production Cost = $625/unit

The company supplies three wholesalers with the following monthly demands:

Wholesaler 1: Demand = 80 units

Wholesaler 2: Demand = 60 units

Wholesaler 3: Demand = 70 units

The unit shipping costs from each plant to each wholesaler are given in the table below.

# 2. Defining Decision Variables
Let Xij represent the number of AEDs shipped from Plant i to Wholesaler j, where:

i∈{A,B} (Plants)
j∈{1,2,3} (Wholesalers)

Therefore, we have 6 decision variables:

xA1,xA2,xA3
xB1,xB2,xB3

# 3. Formulating the Objective Function
The goal is to minimize the total cost, which is the sum of production costs and shipping costs.

Total Cost = Total Production Cost + Total Shipping Cost

We can calculate the cost associated with each route by adding the unit production cost at a plant to the unit shipping cost from that plant to a specific wholesaler. This gives us the total cost per unit for each route.

From \\ To	 Wholesaler1	Wholesaler2	Wholesaler3	 Production Cost
Plant A	    r600+22	    r600+14	        r600+30	    $600
Plant B	    r625+16     r625+20	        r625+24	    $625
Demand	      80	         60	            70	

The objective function coefficients (cost per unit for each route) are:

Plant A to Wholesaler 1: 600+22=622
Plant A to Wholesaler 2: 600+14=614
Plant A to Wholesaler 3: 600+30=630
Plant B to Wholesaler 1: 625+16=641
Plant B to Wholesaler 2: 625+20=645
Plant B to Wholesaler 3: 625+24=649

Thus, the objective function to minimize is:

Min Z = 622xA1+614xA2+630xA3+641xB1+645xB2+649xB3

# Formulating the Constraints
We have two types of constraints: supply constraints and demand constraints.
Supply Constraints (Capacity): The total units shipped from a plant cannot exceed its production capacity.

xA1+xA2+xA3≤100 (Plant A Capacity)
xB1+xB2+xB3≤120(Plant B Capacity)

Demand Constraints: The total units received by a wholesaler must meet its monthly demand.

xA1+xB1=80(Wholesaler 1 Demand)
xA2+xB2=60(Wholesaler 2 Demand)
xA3+xB3=70(Wholesaler 3 Demand)

Non-negativity Constraints: The number of units shipped cannot be negative.

xij ≥ 0 for all i,j

# Solution using Linear Programming in R - We will use the lpSolve package in R to solve this linear programming problem.

```{r}
# The cost vector contains the coefficients of the objective function.
# The order of variables is: x_A1, x_A2, x_A3, x_B1, x_B2, x_B3
cost_coefficients <- c(622, 614, 630, 641, 645, 649)
```

# Step 2: Set up the Constraint Matrix

```{r}
# Create the constraint matrix
constraint_matrix <- matrix(c(
  # Supply Constraints (Rows 1 & 2)
  1, 1, 1, 0, 0, 0,  # Plant A: x_A1 + x_A2 + x_A3 <= 100
  0, 0, 0, 1, 1, 1,  # Plant B: x_B1 + x_B2 + x_B3 <= 120
  # Demand Constraints (Rows 3, 4, 5)
  1, 0, 0, 1, 0, 0,  # Wholesaler 1: x_A1 + x_B1 = 80
  0, 1, 0, 0, 1, 0,  # Wholesaler 2: x_A2 + x_B2 = 60
  0, 0, 1, 0, 0, 1   # Wholesaler 3: x_A3 + x_B3 = 70
), nrow = 5, byrow = TRUE)
```

# Step 3: Set up the Direction and Right-Hand Side (RHS) of Constraints

```{r}
# Define the direction of constraints (<=, =, etc.)
constraint_direction <- c("<=", "<=", "=", "=", "=")

# Define the right-hand side (RHS) values of the constraints
constraint_rhs <- c(100, 120, 80, 60, 70)
```

# Solve the Linear Program

```{r}
# Solve the linear programming problem
lp_solution <- lp("min", cost_coefficients, constraint_matrix, constraint_direction, constraint_rhs)

# Check the status of the solution (0 means success)
lp_solution$status
```

# Display the Optimal Solution

```{r}
optimal_shipments <- lp_solution$solution
names(optimal_shipments) <- c("x_A1", "x_A2", "x_A3", "x_B1", "x_B2", "x_B3")

total_min_cost <- lp_solution$objval

shipping_plan <- matrix(optimal_shipments, nrow = 2, byrow = TRUE,
                        dimnames = list(c("Plant A", "Plant B"),
                                        c("Wholesaler 1", "Wholesaler 2", "Wholesaler 3")))

cat("### Optimal Shipping Plan (Number of Units to Ship)\n")
kable(shipping_plan, format = "simple")

cat("\n### Detailed Variable Values\n")
kable(data.frame(Variable = names(optimal_shipments), Value = optimal_shipments), format = "simple")

cat("\n### Total Minimum Cost\n")
cat("$", format(total_min_cost, big.mark = ",", nsmall = 2), sep = "")
```

# Interpretation of Results

```{r}
plant_a_total <- sum(optimal_shipments[1:3])
plant_b_total <- sum(optimal_shipments[4:6])

wholesaler_totals <- colSums(shipping_plan)

cat("### Total Units Shipped from Each Plant\n")
kable(data.frame(
  Plant = c("Plant A", "Plant B"),
  Total_Shipped = c(plant_a_total, plant_b_total)
))

cat("\n### Total Units Received by Each Wholesaler\n")
kable(data.frame(
  Wholesaler = c("Wholesaler 1", "Wholesaler 2", "Wholesaler 3"),
  Total_Received = wholesaler_totals
))

cat("\n###Interpretation\n")
cat("
The optimal solution indicates how many units each plant should ship to each wholesaler
to achieve the lowest total cost. 

- *Plant A and Plant B* are both fully or efficiently utilized according to their capacities.  
- Each wholesaler’s demand is completely satisfied.  
- The total minimum transportation + production cost is calculated as shown above.  
- This optimal shipping plan ensures the company minimizes logistics expenses while meeting all customer requirements.

In business terms, Heart Start can use this allocation strategy to reduce unnecessary
shipping expenses and balance plant workloads efficiently.
")
```







