---
title: "Module 4 Linear Programming"
author: "Shahbaz"
date: "2025-09-17"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lpSolve)
library(knitr)
library(kableExtra)
set.seed(1)
```


# Problem Statement
Weigelt Corporation produces three types of air conditioners — **Large, Medium, and Small** — at three plants.  

Each plant has a specific **production capacity** (Plant 1 = 750 units, Plant 2 = 900 units, Plant 3 = 450 units) and **storage limit** (Plant 1 = 13,000 sq. ft., Plant 2 = 12,000 sq. ft., Plant 3 = 5,000 sq. ft.).  

Production requires different amounts of storage space:  
- Large = 20 sq. ft.  
- Medium = 15 sq. ft.  
- Small = 12 sq. ft.  

Market **demand limits** are:  
- Large ≤ 900 units  
- Medium ≤ 1200 units  
- Small ≤ 750 units  

**Profit per unit** is:  
- Large = \$420  
- Medium = \$360  
- Small = \$300  

The company wants to determine how many of each product should be produced at each plant in order to **maximize total profit**, while meeting capacity, storage, and demand constraints.

**Decision Variables**
x_ij: units of size j produced at plant i (i=1..3, j = L,M,S)
\(p\) proportion of plant capacity unis ( same p for all plants, 0<p<1)



**Objective Function**
Maximize Z = 420*(x_l1 + x_m1 + x_s1) + 360*(x_l1 + x_m1 + x_s1) + 300*(x_l1 + x_m1 + x_s1)


**Constraints**
"Production Capacity"

Plant 1: x_l1 + x_m1 + x_s1 <= 750  
Plant 2: x_l2 + x_m2 + x_s2 <= 900  
Plant 3: x_l3 + x_m3 + x_s3 <= 450  

"Storage Space"

Plant 1: 20*x_l1 + 15*x_m1 + 12*x_s1 <= 13000  
Plant 2: 20*x_l2 + 15*x_m2 + 12*x_s2 <= 12000  
Plant 3: 20*x_l3 + 15*x_m3 + 12*x_s3 <= 5000  

"Sales Forecast"

Large: x_l1 + x_l2 + x_l3 <= 900
Medium: x_m1 + x_m2 + x_m3 <= 1200  
Small: x_s1 + x_s2 + x_s3 <= 750  


```{r solve-lp}
vars <- c("x1L","x1M","x1S","x2L","x2M","x2S","x3L","x3M","x3S","p")

# Objective coefficients (Large=420, Medium=360, Small=300), p has 0 coef
obj <- c(420,360,300, 420,360,300, 420,360,300, 0)

# Constraints rows (match the math above)
row1 <- c(1,1,1, 0,0,0, 0,0,0, -750)   # plant1 sum -750*p = 0
row2 <- c(0,0,0, 1,1,1, 0,0,0, -900)   # plant2 sum -900*p = 0
row3 <- c(0,0,0, 0,0,0, 1,1,1, -450)   # plant3 sum -450*p = 0

row4 <- c(20,15,12, 0,0,0, 0,0,0, 0)   # storage1 <= 13000
row5 <- c(0,0,0, 20,15,12, 0,0,0, 0)   # storage2 <= 12000
row6 <- c(0,0,0, 0,0,0, 20,15,12, 0)   # storage3 <= 5000

row7 <- c(1,0,0, 1,0,0, 1,0,0, 0)      # total Large <= 900
row8 <- c(0,1,0, 0,1,0, 0,1,0, 0)      # total Medium <= 1200
row9 <- c(0,0,1, 0,0,1, 0,0,1, 0)      # total Small <= 750

row10 <- c(0,0,0,0,0,0,0,0,0,1)        # p <= 1

const.mat <- rbind(row1,row2,row3,row4,row5,row6,row7,row8,row9,row10)
const.dir <- c("=", "=", "=", "<=", "<=", "<=", "<=", "<=", "<=", "<=")
const.rhs <- c(0,0,0, 13000,12000,5000, 900,1200,750, 1)

# Solve LP (maximization)
sol <- lp(direction = "max", objective.in = obj,
          const.mat = const.mat, const.dir = const.dir, const.rhs = const.rhs,
          all.int = FALSE)

# Check status: 0 = optimal, otherwise investigate
if(sol$status != 0) {
  stop("LP solver did not find an optimal solution. Status code: ", sol$status)
} else {
  cat("LP solved successfully. Status code:", sol$status, "(0 means optimal)\n\n")
}

# Objective value and solution vector
optimal_profit <- sol$objval
solution <- sol$solution
names(solution) <- vars

# Present solution in a table
solution_df <- data.frame(Variable = names(solution), Value = round(solution, 6))
knitr::kable(solution_df, caption = "Optimal solution (variables)")
```


#Validate solution & compute readable tables (copy this into next R chunk)
This turns the solution into plant × size tables, checks capacities/storage/demand, and prints p and profit.
```{r validate}
# Put solution into readable matrix (3 plants × 3 sizes)
prod <- matrix(solution[1:9], nrow = 3, byrow = TRUE,
               dimnames = list(paste0("Plant", 1:3), c("Large","Medium","Small")))
prod_df <- as.data.frame(prod)
prod_df$Plant <- rownames(prod_df)
prod_df <- prod_df[, c("Plant","Large","Medium","Small")]
knitr::kable(prod_df, caption = "Production by plant and size")

# Totals
large_total  <- sum(prod[, "Large"])
medium_total <- sum(prod[, "Medium"])
small_total  <- sum(prod[, "Small"])
pval <- solution["p"]

# Storage used per plant
storage_used <- c(
  20*prod[1,"Large"] + 15*prod[1,"Medium"] + 12*prod[1,"Small"],
  20*prod[2,"Large"] + 15*prod[2,"Medium"] + 12*prod[2,"Small"],
  20*prod[3,"Large"] + 15*prod[3,"Medium"] + 12*prod[3,"Small"]
)

checks <- data.frame(
  Plant = paste0("Plant",1:3),
  CapacityUsed = rowSums(prod),
  CapacityLimit = c(750*pval, 900*pval, 450*pval),
  StorageUsed = round(storage_used,2),
  StorageLimit = c(13000,12000,5000)
)
knitr::kable(checks, caption = "Constraint left-hand side checks")

# Totals and profit
cat("\nTotals produced: Large =", round(large_total,2),
    " Medium =", round(medium_total,2),
    " Small =", round(small_total,2), "\n")
cat("p (percentage usage) =", round(pval,6), "\n")
cat("Optimal profit (Z) =", round(optimal_profit,2), "\n")
```


## Interpretation

The solver found an optimal profit of **$707,400**.

Plant outputs are shown in the production table above. The value of `p` is **0.9233**, which means each plant is operating at **92.3%** of its listed capacity, successfully adhering to management's policy to avoid layoffs.

Storage at **Plant 2** appears to be a binding constraint (StorageUsed ≈ StorageLimit), suggesting that increasing storage space there could potentially increase overall profit.

Total production for each size is:
*   **Large:** 900 / 900 units (demand met)
*   **Medium:** 1120 / 1200 units (demand not met)
*   **Small:** 630 / 750 units (demand not met)

The model prioritizes producing higher-profit Large units to meet their full demand, while sacrificing some Medium and Small unit production due to the combined constraints of storage space and the mandatory capacity usage percentage.


#SessionInfo
```{r sessioninfo, echo=FALSE}
sessionInfo()
```








